---
title: "Evaluating Numerical Fidelity in Financial Documents: A Comparative Analysis of Naive and Table-Aware Dual-Branch RAG Architectures"
subtitle: "A Capstone Project in Data Science"
author(s): "Crystal Shin, Gredy Garrido, and Nick Lopez"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

Slides: [slides.html](slides.html){target="_blank"} ( Go to `slides.qmd`
to edit)

## Introduction

### The Real-World Risk: Why Tiny AI Mistakes Matter
Today, many companies use AI assistants to quickly find information buried in their internal reports. These systems are great at summarizing emails or articles, but they often fail when reading financial tables [@guo2026ftrag]. Imagine an employee asking the AI system for a company’s total debt from a dense financial document. In a typical financial table, a number like “450” might mean $450 million if it is listed under a header that says “figures in millionsthousands.” Because standard AI reads text in a straight line from left to right, it can lose track of those important headers and labels. It might report the debt as just $450 or link the number to the wrong year entirely. In the world of finance, a mistake like this isn’t just a small typo; it can lead to massive losses or legal trouble. 

This problem is commonly known as the gap between what a table says and what the AI actually understands. The dominant approach for building document-querying AI systems is Retrieval-Augmented Generation (RAG) [@malali2025role] — a method that retrieves relevant passages from a document library before generating a response, grounding the AI’s answer in real source material. While RAG has proven effective for regulatory compliance, audit reporting, and other text-heavy financial tasks, its standard pipeline was designed for narrative prose and breaks down when applied to structured tables. This paper compares a standard RAG baseline against a table-aware pipeline on a corpus of real financial reports, introducing a Numerical Atomic Verification (NAV) metric to explicitly measure numerical fidelity in model outputs.

### The Core Problem Statement: The Structural Evidence Gap

The core issue is that standard AI systems struggle to read complex financial tables. While these systems are great at reading plain text, they fail when processing financial data because they try to parse tables line by line, as if they were sentences [@guo2026ftrag]. We refer to this as the Structural Evidence Gap. This leads to two main problems:

  1. **Context Loss**: When the AI system breaks text into smaller chunks, it separates numbers from their labels (like headers or rows). This means the AI loses track of what the numbers actually represent.
  2. **Formatting Issues**: Because standard tools read tables linearly, they jumble the data, mixing up different rows and columns.

While these AI systems work well for general reading, they fail when analyzing complex financial records in which answers are hidden across multiple structured tables.

### Overview of Approach: Table-Aware Dual-Branch RAG

To solve the problems described above, this paper compares two approaches:

  - Architecture A (Naive RAG): This standard method uses simple linear PDF reading and basic semantic search, which often misses the structure of financial tables.
  - Architecture B (Table-Aware Dual-Branch RAG): This is our proposed solution, which helps retain structural lineage between document context and table information.

Our approach includes three key steps:

  1. Bifurcated Ingestion: The system separates narrative text from tabular data at the beginning, ensuring tables are preserved as structured data rather than flattened into plain text.
  2. Hybrid Indexing: We store each table row as an individual record, enriched with its column/row labels [@table_aware_chinese]. This ensures the AI system always knows exactly what a number represents (e.g., Fiscal Year 2025 vs. “2025”).
  3. Numerical Atomic Verification (NAV): Before the system provides an answer, a final verification step automatically checks all numbers against the original source document. If a discrepancy or unit mismatch is found, it is flagged to ensure absolute mathematical accuracy.

The pipeline is benchmarked against standard industry metrics to prove its effectiveness in handling complex financial documents.

## Methods

-   Detail the models or algorithms used.

-   Justify your choices based on the problem and data.

*The common non-parametric regression model is*
$Y_i = m(X_i) + \varepsilon_i$*, where* $Y_i$ *can be defined as the sum
of the regression function value* $m(x)$ *for* $X_i$*. Here* $m(x)$ *is
unknown and* $\varepsilon_i$ *some errors. With the help of this
definition, we can create the estimation for local averaging i.e.*
$m(x)$ *can be estimated with the product of* $Y_i$ *average and* $X_i$
*is near to* $x$*. In other words, this means that we are discovering
the line through the data points with the help of surrounding data
points. The estimation formula is printed below [@R-base]:*

$$
M_n(x) = \sum_{i=1}^{n} W_n (X_i) Y_i  \tag{1}
$$$W_n(x)$ *is the sum of weights that belongs to all real numbers.
Weights are positive numbers and small if* $X_i$ *is far from* $x$*.*

*Another equation:*

$$
y_i = \beta_0 + \beta_1 X_1 +\varepsilon_i
$$

## Analysis and Results

### Data Exploration and Visualization

-   Describe your data sources and collection process.

-   Present initial findings and insights through visualizations.

-   Highlight unexpected patterns or anomalies.

A study was conducted to determine how...


```{r, warning=FALSE, echo=T, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
```
```{python}
import pandas as pd

```

```{r, warning=FALSE, echo=TRUE}


# Load Data
kable(head(murders))

ggplot1 = murders %>% ggplot(mapping = aes(x=population/10^6, y=total)) 

  ggplot1 + geom_point(aes(col=region), size = 4) +
  geom_text_repel(aes(label=abb)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(formula = "y~x", method=lm,se = F)+
  xlab("Populations in millions (log10 scale)") + 
  ylab("Total number of murders (log10 scale)") +
  ggtitle("US Gun Murders in 2010") +
  scale_color_discrete(name = "Region")+
  theme_bw()


```

### Modeling and Results

-   Explain your data preprocessing and cleaning steps.

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**

```{r}

```

### Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.

## References
